---
title: "Day 1 Part 2 Exercise Michael Scheriger"
date: "`r Sys.Date()`"
output: html_document

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Load the Gapminder Data

```{r Load Data}
# Here is where your R code goes. ``` opens the code chunk, followed by {r}. Then ``` closes the chunk. When you 'knit' the file, the codes will evaluate.

gapminder <- read.csv(here::here("data/gapminder5.csv"), stringsAsFactors=FALSE)
```

##Check to see if there is data from a given year

```{r Check year}
years <- unique(gapminder$year)
for (year in c(2002,2012)){
  if (any(years == year)){
    print(paste0("There is data from the year ", year))
  }
  else{
    print(paste0("There is not data from the year ", year))
  }
}
```

##Mean life expectancy for countries with below average population.

```{r Mean life expectancy}
avg_pop = mean(gapminder$pop)
for (c in unique(gapminder$country)){
  c_pop = mean(gapminder$pop[gapminder$country == c])
  if(c_pop < avg_pop){
    mean_le <- mean(gapminder$lifeExp[gapminder$country == c])
    print(paste0("The average life expectancy for ",c," is: ",mean_le))
  }
}
```

##Life Expectancy by Continent
Write a script that loops through the `gapminder` data by continent and prints out whether the mean life expectancy is smaller than 50, between 50 and 70, or greater than 70.

```{r Life Expectancy by Continent}
for (c in unique(gapminder$continent)){
  if(mean(gapminder$lifeExp[gapminder$continent==c]) < 50){
    print(paste0(c," has a life expectancy smaller than 50"))
  }
  else if(mean(gapminder$lifeExp[gapminder$continent==c]) > 70){
    print(paste0(c," has a life expectancy greater than 70"))
  }
  else{
    print(paste0(c," has a life expectancy between 50 and 70"))
  }
}
```

##Print column name and class
Create a function that given a data frame will print the name of each column and the class of data it contains. Use the gapminder dataset.  Hint: Use `mode()` or `class()` to get the class of the data in each column. Remember that `names()` or `colnames()` returns the name of the columns in a dataset.

```{r Print column name and class}
print_name_class <- function(df){
  name <- names(df)
  for (n in name){
    print(paste0("Column name is ",n, " and the class is ",class(df[[n]])))
  }
}
```

##Return mean, sd, and (maybe) median
Create a function that given a vector will print the mean and the standard deviation of a **vector**, it will optionally also print the median. Hint: include an argument that takes a boolean (`TRUE`/`FALSE`) operator and then include an `if` statement.

```{r Return stats}

stat_return <- function(vec, med = TRUE){
  print(paste0("Mean of the vector is: ", mean(vec)))
  print(paste0("Standard Deviation of the vector is: ", sd(vec)))
  if (med == TRUE){
    print(paste0("Median of the vector is: ", median(vec)))
  }
}
```

##Analyze GDP Per Capita and Life Expectancy
1. What is the relationship between GDP per capita and life expectancy? Does this relationship change over time? (Hint: Use the natural log of both variables.)

```{r GDP v LE}
for (year in unique(gapminder$year)){
  g <- gapminder[gapminder$year == year,]
  plot(log(g$gdpPercap),log(g$lifeExp), main = paste0("Log GDP Per Cap vs. Log Life Expectancy in ",year),xlab = "Log GDP Per Cap", ylab = "Log Life Expectancy")
  abline(lm(log(g$lifeExp) ~ log(g$gdpPercap)), col="red")
}
```

##Analyze GDP per capita and life expectancy by Continent
Does the relationship between GDP per capita and life expectacy vary by continent? Make sure you divide the Americas into North and South America.

```{r By Continent}
for (c in unique(gapminder$continent)){
  g <- gapminder[gapminder$continent == c,]
  plot(log(g$gdpPercap),log(g$lifeExp), main = paste0("Log GDP Per Cap vs. Log Life Expectancy in ",c),xlab = "Log GDP Per Cap", ylab = "Log Life Expectancy")
  abline(lm(log(g$lifeExp) ~ log(g$gdpPercap)), col="red")
}
```